%   Filename    : chapter_4.tex 
\chapter{Research Methodology}
This chapter lists and discusses the specific steps and activities that will be performed to accomplish the project. 
The discussion covers the activities from pre-proposal to Final SP Writing.

\section{Research Activities}

\subsection{Gather Dengue Data and Climate Data to Create a Complete Dataset for Forecasting}

\subsubsection{Acquisition of Dengue Case Data}
\begin{itemize}
	\item Request and secure historical dengue case data for the Western Visayas region from:
	\begin{itemize}
		\item Humanitarian Data Exchange.
		\item Western Visayas Center for Health Development (WVCHD) via Freedom of Information (FOI) requests.
	\end{itemize}
	\item Validate data accuracy and completeness to ensure reliability.
\end{itemize}

\subsubsection{Collection of Weather Data}
\begin{itemize}
	\item Use web scraping techniques to extract weekly weather data (e.g., rainfall, temperature, and humidity) from Weather Underground.
\end{itemize}

\subsubsection{Data Integration and Preprocessing}
\begin{itemize}
	\item Combine dengue case data with weather data into a complete dataset that aligns data based on timeframes.
	\item Clean data to address missing values, outliers, and inconsistencies.
\end{itemize}

\subsubsection{Exploratory Data Analysis (EDA)}
\begin{itemize}
	\item Analyze trends, seasonality, and correlations between dengue cases and weather factors.
	\item Create visualizations like time series plots and scatterplots to highlight relationships and patterns in the data.
\end{itemize}

\subsection{Develop and Evaluate Deep Learning Models for Dengue Case Forecasting}

\subsubsection{Algorithm Selection for Time Series Modeling}
\begin{itemize}
	\item Evaluate the suitability of algorithms for time series forecasting, focusing on:
	\begin{itemize}
		\item Deep learning models like LSTM (Long Short-Term Memory).
		\item Traditional statistical models ARIMA and Seasonal ARIMA.
		\item State-space models Kalman Filter.
	\end{itemize}
	\item Assess each algorithm's strengths in handling temporal patterns and external factors.
\end{itemize}

\subsubsection{Model Training and Testing}
\begin{itemize}
	\item Split the dataset into training and testing sets while maintaining temporal continuity.
	\item Train each model using historical data and tune hyperparameters to enhance accuracy.
	\item Evaluate models using performance metrics such as MSE and RMSE.
\end{itemize}

\subsubsection{Model Evaluation and Optimization}
\begin{itemize}
	\item Compare the performance of all models to identify the most accurate forecasting approach.
	\item Iteratively optimize the selected model.
\end{itemize}

\subsection{Integrate the Predictive Model into a Web-Based Data Analytics Dashboard}

\subsubsection{Dashboard Design and Development}
\begin{itemize}
	\item Design an intuitive, user-friendly web-based dashboard incorporating:
	\begin{itemize}
		\item Interactive visualizations of yearly dengue case trends.
		\item Data input and update forms for dengue and weather data.
		\item Map display of dengue cases in each district in Iloilo City
	\end{itemize}
\end{itemize}

\subsubsection{Model Integration and Deployment}
\begin{itemize}
	\item Deploy the best-performing model within the dashboard as a backend service to enable real-time or periodic forecasting.
\end{itemize}


\subsection{System Development Framework}
The Agile Model is the birthchild of both iterative and incremental approaches in Software Engineering. It aims to be flexible and effective at the same time by being adaptable to change. It's also important to note that small teams looking to construct and develop projects quickly can benefit from this kind of methodology. As the Agile Method focuses on continuous testing, quality assurance is a guarantee since bugs and errors are quickly identified and patched. 

\subsection{Design, Building, Testing, and Integration}
\subsubsection{Design and Developlment}
After brainstorming and researching the most appropriate type of application to accommodate both the prospected users and the proposed solutions, the team has decided to proceed with a web application. Given the time constraints and available resources, we believe this is the most pragmatic and practical move. The next step is to select modern and stable frameworks that align with the fundamental ideas we have learned at the university. The template obtained from WVCHD was meticulously analyzed to create use cases and develop a preliminary well-structured database that adheres to the requirements needed to produce a quality application. The said use cases serve as the basis of general features. Part by part, these are converted into code, and with the help of selected libraries and packages, it resulted in the desired outcome that may still modified and extended since it is continuously being developed. 

\subsubsection{Testing and Integration}
Each feature is rigorously user-tested to ensure quality assurance, with particular emphasis on prerequisite features, as development cannot progress properly if these fail. Moreover, integration between each feature serves as a pillar for a cohesive user experience. Presently, we have not been able to use performance metrics to measure the system's performance, as developing and connecting the core features is the utmost priority. 

\section{Model and Algorithm}
\textbf{Data Preprocessing and Feature Selection:}
\begin{itemize}
	\item Clean and preprocess the dataset, handling missing values and outliers.
	\item Select important features such as temperature, rainfall, and humidity.
\end{itemize}

\textbf{Model Selection and Justification:}
\begin{itemize}
	\item Train an LSTM model due to its efficiency in handling long-term data relationships.
	\item Compare performance with ARIMA, SARIMA, and Kalman Filter.
\end{itemize}

\textbf{Seasonal ARIMA (SARIMA):}
\begin{itemize}
	\item Set seasonal parameters based on observed seasonality in cases.
	\item Apply SARIMA to capture short-term and seasonal trends.
\end{itemize}

\textbf{Kalman Filter:}
\begin{itemize}
	\item Use Kalman Filter for real-time updates and handle missing values dynamically.
\end{itemize}

\textbf{LSTM (Long Short-Term Memory) Neural Network:}
\begin{itemize}
	\item Build a sequential LSTM model with tuned hyperparameters.
\end{itemize}

\textbf{Model Evaluation:}
\begin{itemize}
	\item Evaluate performance with Mean Absolute Error (MAE), RMSE, and R-squared metrics.
\end{itemize}

\section{Development Tools}
\subsection{Software}

\subsubsection{Github}
GitHub is a cloud-based platform that tracks file changes using Git, an open-source version control system \cite{github-no-date}. It is used in the project to store the application's source code, manage the system's source version control, and serve as a repository for the Latex files used in the actual research.

\subsubsection{Visual Studio Code}
Visual Studio Code is a free, lightweight, and cross-platform source code editor developed by Microsoft \cite{vscode-2021}. As VS Code supports this project's programming and scripting languages, it was chosen as the primary source code editor.

\subsubsection{Django}
Django is a free and open-sourced Python-based web framework that offers an abstraction to develop and maintain a secure web application. As this research aims to create a well-developed and maintainable application, it is in the best interest to follow an architectural pattern that developers and contributors in the future can understand. Since Django adheres to Model-View-Template (MVT) that promotes a clean codebase by separating data models, business logic, and presentation layers, it became the primary candidate for the application's backbone. 


\subsubsection{Next.js}
A report by Statista (2024) claims that React is the most popular front-end framework among web developers. However, React has limitations that can be a nuisance in rapid software development, which includes routing and performance optimizations. This is where Next.js comes in—a framework built on top of React. It offers solutions for React's deficiency, making it a rising star in the framework race. 

\subsubsection{Postman}
As the application heavily relies on the Application Programming Interface (API) being thrown by the backend, it is a must to use a development tool that facilitates the development and testing of the API. Postman is a freemium API platform that offers a user-friendly interface to create and manage API requests \cite{postman-no-date}. 

\subsection{Hardware}
The web application is continuously being developed on laptop computers with minimum specifications of an 11th-generation Intel i5 CPU and 16 gigabytes of RAM.

\subsection{Packages}

\subsubsection{Django REST Framework}
Django Rest Framework (DRF) is a third-party package for Django that provides a comprehensive suite of features to simplify the development of robust and scalable Web APIs \cite{christie-no-date}. These services include Serialization, Authentication and Permissions, Viewsets and Routers, and a Browsable API . 

\subsubsection{Leaflet}
One of the features of the web application is the ability to map the number of cases using a Choropleth Map. Leaflet is the only free, open-sourced, and most importantly, stable JavaScript package that can do the job. With its ultra-lightweight size, it offers a comprehensive set of features that does not trade off performance and usability \cite{leaflet-no-date}. 

\subsubsection{Chart.js}
Another feature of the application is to provide users with informative, approachable data storytelling that is easy for everyone to understand. The transformation of pure data points and statistics into figures such as charts is a big factor. Thus, there is a need for a package that can handle this feature without compromising the performance of the application. Chart.js is a free and open-source JavaScript package that is made to meet this criteria as it supports various types of charts \cite{chartjs-no-date}. 

\subsubsection{Tailwind CSS}
Using plain CSS in production-quality applications can be counterproductive. Therefore, CSS frameworks were developed to promote consistency and accelerate the rapid development of web applications \cite{joel-2021}. One of these is Tailwind, which offers low-level utility classes that can be applied directly to each HTML element to create a custom design \cite{tailwind-no-date}. Given the limited timeline for this project, using this framework is a wise choice due to its stability and popularity among developers.

\subsubsection{Shadcn}
Shadcn offers a collection of open-source UI boilerplate components that can be directly copied and pasted into one's project. With the flexibility of the provided components, Shadcn allows developers to have full control over customization and styling. Since this is built on top of Tailwind CSS and Radix UI, it is supported by most modern frontend frameworks, including Next.js \cite{shadcn-no-date}.

\section{Calendar of Activities}

A Gantt chart showing the schedule of the activities is included below. Each bullet represents approximately one week of activity.

\newcommand{\weekone}{\textbullet}
\newcommand{\weektwo}{\textbullet \textbullet}
\newcommand{\weekthree}{\textbullet \textbullet \textbullet}
\newcommand{\weekfour}{\textbullet \textbullet \textbullet \textbullet}

\begin{table}[ht]
	\centering
	\caption{Timetable of Activities} \vspace{0.25em}
	\begin{tabular}{|p{2in}|c|c|c|c|c|c|c|c|} \hline
		Activities (2024) & Aug & Sept & Oct & Nov & Dec \\ \hline
		Project Initiation and Team Formation & \weektwo & & & & \\ \hline
		Literature Review and Data Gathering & \weektwo & \weekfour & & & \\ \hline
		Data Cleaning and Feature Selection & & \weektwo & & \weekone & \\ \hline
		
		
		Creating System Dashboard & & \weektwo & \weekfour & \weekone & \\ \hline
		Analysis and Interpretation of Results & & & \weekone & & \\ \hline
		Documentation & & \weekone & \weekfour & \weekone & \\ \hline
	\end{tabular}
	\label{tab:timetableactivities}
\end{table}
